A highly-available move operation for replicated trees

### 2019-06-20 paper/move-op.tex
paper: Section --> \S throughout
Some other minor save-spacing edits here and there.

Many systems use a tree-structured data model.
For example:
\begin{itemize}
    \item The \textbf{filesystem} on most operating systems is organised as a tree: folders or directories are branch nodes, and files are leaf nodes.
        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in Section~\ref{sec:hardlinks} how to add support for links.
    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
        These objects are contained within nodes representing pages of a document, or slides of a presentation.
        Multiple objects may be combined into a \emph{group} so that they can be manipulated as a unit; this corresponds to making these objects children of a common parent node.
        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
    \item \textbf{Note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} present the user with a tree structure that they can inspect and manipulate.
\end{itemize}

@@

In this paper we use an abstract model of a tree that can support any of the applications listed above, including distributed filesystems.
We allow replicas to manipulate this tree in any way: by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
While there are many existing systems that support creating and deleting nodes (see Section~\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
We explain in Section~\ref{sec:move-is-hard} why this move operation is so challenging.
In summary, our contributions are:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
        As discussed in Section~\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.
        We then extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions and evaluate its performance with replicas across three continents.
    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

Applications that rely on a tree data model often need to move a node from one location to another location within the tree, whereby all children of the moved node move along with it.
For example:
\begin{itemize}
    \item In a filesystem, any file or directory can be moved to a different parent directory.
        Moreover, renaming a file or directory is equivalent to moving it to a new name without changing its parent directory.
    \item In a rich text editor, a paragraph can be turned into a bullet point.
        In the DOM tree this corresponds to creating a new bullet list node, creating a new bullet point node within it, and then moving the paragraph node inside the bullet point.
    \item In presentation software, grouping two graphical objects corresponds to creating a new group node, and then moving the two objects to be children of the new group node.
\end{itemize}
As these operations are so common, it is not obvious why a move operation should be difficult in a replicated setting.
In this section we demonstrate by example some of the problems that arise with replicated trees, before proceeding to our solution in Section~\ref{sec:algorithm}.

@@

Another possible resolution is shown in Figure~\ref{fig:move-cycle}b: the nodes involved in the concurrent moves (and their children) could be duplicated, so that both ``$A$ as a child of $B$'' and ``$B$ as a child of $A$'' can exist in the tree.
However, such duplication is undesirable for the same reasons as in Section~\ref{sec:move-same-item}.

In our opinion, the best way of handling the conflicting operations of Figure~\ref{fig:move-cycle} is to choose either Figure~\ref{fig:move-cycle}c or~\ref{fig:move-cycle}d: that is, either the result of applying replica 1's operation and ignoring replica 2's operation, or vice versa.
Like in Section~\ref{sec:move-same-item}, the winning operation can be picked based on an operation timestamp.

@@

Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} have previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle, as in Section~\ref{sec:move-cycle}.
Using the CISE ('Cause I'm Strong Enough) proof tool~\cite{DBLP:conf/popl/GotsmanYFNS16,Najafzadeh:2016fi} the authors confirm that it is not sufficient for the replica that generates a move operation to check whether the operation introduces a cycle: like in Figure~\ref{fig:move-cycle}, two concurrent operations may be safe individually, but introduce a cycle when combined.

@@

While these solutions are valid, the authors go on to claim that ``no file system can support an unsynchronised move without anomalies, such as loss or duplication''~\cite{Najafzadeh:2018bw}.
We revisit that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisified (in particular, it never introduces cycles), and it never duplicates or loses any tree nodes.
To our knowledge, our algorithm is the first to provide all of these properties simultaneously.
We give a precise specification of our algorithm's consistency properties in Section~\ref{sec:proof}.

@@

The key consistency property of our algorithm is \emph{convergence}: that is, whenever any two replicas have applied the same set of operations, then they must be in the same state---even if the operations were applied in a different order on different replicas.
We prove this in Section~\ref{sec:proof} by showing that applying operations is commutative.

Figure~\ref{fig:code} gives the full source code for our algorithm, implemented using the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
We choose this language because it combines the conciseness of pseudocode with the precision of mathematical notation.
It supports formal reasoning, allowing us to prove the correctness of the algorithm (Section~\ref{sec:proof}), and it also allows executable code to be exported to common functional programming languages such as Scala, Haskell, or OCaml.

@@

Recall from Section~\ref{sec:move-cycle} that in order to prevent a cycle being introduced, the node being moved must not be an ancestor of the destination node.
To implement this in Figure~\ref{fig:code}, we first define the \isa{ancestor} relation in lines 22--24.
This relation is the transitive closure of the parent-child relation of a tree: if \isa{(p, m, c) {\isasymin} tree} then \isa{p} is an ancestor of \isa{c} (line 23); moreover, if \isa{(p, m, c) {\isasymin} tree} and \isa{a} is an ancestor of \isa{p}, then \isa{a} is also an ancestor of \isa{c} (line 24).
The \textbf{inductive} keyword indicates that this recursive definition is iterated until the least fixed point is reached.

The \isa{do\_op} function (lines 26--31) now performs the actual work of applying a move operation.
This function takes as argument a pair consisting of a \isa{Move} operation and the current tree, and it returns a pair consisting of a \isa{LogMove} operation (which will be added to the log) and an updated tree.
In line 29, the \isa{LogMove} record is constructed as described in Section~\ref{sec:state-log}, obtaining the prior parent and metadata of \isa{c} using the \isa{get\_parent} function.

@@

Due to the undo-do-redo cycle, the state of the tree is as if all operations had been applied using \isa{do\_op} in increasing timestamp order, regardless of the order in which they were actually applied.
This provides a clear approach to the handling of conflicts:
\begin{itemize}
    \item If two operations concurrently move the same node, then the ``winning'' operation is the one with the higher timestamp.
        Effectively, the operation with the lower timestamp moves the node first, and then the operation with the higher timestamp moves it again.
        Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
    \item If two operations would introduce a cycle when combined, as in Section~\ref{sec:move-cycle}, then the operation with the greater timestamp is ignored.
        This is the case because \isa{do\_op} checks for cycles based on the tree created by all operations with a lower timestamp.
        The lower of two conflicting operations will take effect, since that operation by itself is safe.
        When the higher-timestamped operation is applied, \isa{do\_op} detects that it would introduce a cycle, and therefore ignores the operation.
\end{itemize}

@@

In a distributed filesystem, replication and conflict resolution is required not only for the directory structure, but also for the contents of individual files.
This can be accomplished by using CRDTs for file contents as well.
For example, we have explored CRDT algorithms for applications using a JSON data model~\cite{Kleppmann:2016ve,LocalFirst} and implemented them in a library called Automerge~\cite{Automerge}.
We discuss distributed filesystems further in Section~\ref{sec:filesystems}.

@@

We now discuss the correctness properties of the algorithm from Section~\ref{sec:algorithm}.
All theorems stated here have been formally proved and mechanically checked using Isabelle.
For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in our open source Isabelle files~\cite{SourceFiles}.

To reason about the state of a replica we define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code}.
It takes a list of operations \isa{ops} and returns the state of a replica after it has applied all the operations in \isa{ops}.
The \isa{apply\_ops} function works by starting in the initial state \isa{([], \{\})} consisting of the empty operation log \isa{[]} and the tree represented by the empty set \isa{\{\}}, and then applying the operations one by one to the state using the \isa{apply\_op} function (introduced in Section~\ref{sec:applying}).
The \isa{foldl} function from the Isabelle/HOL standard library performs the iteration over the list of operations.

@@

In fact, we slightly generalise this property: we allow more than one root to exist, so the graph is actually a forest.
This allows an application to move nodes between multiple trees, if desired.
For example, the ``trash'' node used for deletion (Section~\ref{sec:ops-trees}) can be separate from the main tree.

@@

As discussed in Section~\ref{sec:applying}, we require that when replicas apply the same set of operations, they converge towards the same state, regardless of the order in which the operations are applied.
We formalise this in Isabelle as follows:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}commutes{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}set\ ops{\isadigit{1}}\ {\isacharequal}\ set\ ops{\isadigit{2}}{\isacartoucheclose}\isanewline
\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{1}}{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops{\isadigit{1}}\ {\isacharequal}\ apply{\isacharunderscore}ops\ ops{\isadigit{2}}{\isacartoucheclose}
\end{isabelle}

@@

Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL definitions using a sophisticated code generation mechanism~\cite{DBLP:conf/flops/HaftmannN10}.
However, not all definitions can be realised in executable form: for example, the use of choice principles (like in \isa{get\_parent}) and inductively defined relations (e.g.\ \isa{ancestor}) cause problems.
Moreover, HOL's \isa{set} type allows infinite sets.
Whilst these constructs are convenient for theorem proving, they do not translate well to efficient code.

We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
Rather than representing the tree as a set of (parent, metadata, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (metadata, parent) pairs, written in Isabelle as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
The \isa{hashable} type class on keys means that keys must have an associated hashing function.
In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (Section~\ref{sec:unique-parent}).
We use the hash-map implementation from the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.

@@

That is, if both algorithms are applied to the same set of operations then they produce identical logs and also produce trees that contain the same set of key-value bindings---i.e. the two trees are extensionally equivalent, despite having very different concrete representations.
Moreover, we have the following corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our executable algorithms:\footnote{The ``executable'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:

@@

Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
CRDTs have been defined for various abstract datatypes, such as registers and counters \cite{Shapiro:2011wy,Shapiro:2011un}, maps \cite{Baquero:2016iv,Kleppmann:2016ve}, sets \cite{Bieniusa:2012wu,Bieniusa:2012gt}, lists (sequences) \cite{Roh:2011dw,Nedelec:2013ky}, and text \cite{Preguica:2009fz,Weiss:2010hx}.
For tree data structures, several approaches have been proposed:
\begin{itemize}
    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support a move operation.
    \item As discussed in Section~\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two alternative implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} survey approaches to handling conflicts on trees without providing concrete algorithms.
    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
        Some conflicts are handled by duplicating tree nodes.
        Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in Section~\ref{sec:move-is-hard}.
    \item In our previous unpublished work \cite{ExtendedVersion} we develop the foundations for the algorithm described in this paper.
\end{itemize}

>>>>

Many systems use a tree-structured data model.
For example:
\begin{itemize}
    \item The \textbf{filesystem} on most operating systems is organised as a tree: folders or directories are branch nodes, and files are leaf nodes.
        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in \S\ref{sec:hardlinks} how to add support for links.
    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
        These objects are contained within nodes representing pages of a document, or slides of a presentation.
        Multiple objects may be combined into a \emph{group} so that they can be manipulated as a unit; this corresponds to making these objects children of a common parent node.
        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
    \item \textbf{Note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} present the user with a tree structure that they can inspect and manipulate.
\end{itemize}

@@

In this paper we use an abstract model of a tree that can support any of the applications listed above, including distributed filesystems.
We allow replicas to manipulate this tree in any way: by creating new nodes, deleting nodes, or moving subtrees to a new location within the tree.
While there are many existing systems that support creating and deleting nodes (see \S\ref{sec:relwork}), the key innovation of our algorithm is the support for moving subtrees.
We explain in \S\ref{sec:move-is-hard} why this move operation is so challenging.
In summary, our contributions are:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that supports a \emph{move} operation, without requiring any coordination between replicas such as locking or consensus.
        As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise an abstract algorithm using Isabelle/HOL~\cite{DBLP:conf/tphol/WenzelPN08}, an interactive proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness of the algorithm.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge towards a consistent state (strong eventual consistency) while keeping the data in a tree structure.
    \item To demonstrate the practical viability of our approach, we refine our abstract algorithm to an efficient implementation within Isabelle/HOL and prove the equivalence of the two.
        We then extract a formally verified, executable Scala implementation from the Isabelle/HOL definitions and evaluate its performance with replicas across three continents.
    \item We perform experiments with replicated filesystem products such as Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

Applications that rely on a tree data model often need to move a node from one location to another location within the tree, whereby all children of the moved node move along with it.
For example:
\begin{itemize}
    \item In a filesystem, any file or directory can be moved to a different parent directory.
        Moreover, renaming a file or directory is equivalent to moving it to a new name without changing its parent directory.
    \item In a rich text editor, a paragraph can be turned into a bullet point.
        In the DOM tree this corresponds to creating a new bullet list node, creating a new bullet point node within it, and then moving the paragraph node inside the bullet point.
    \item In presentation software, grouping two graphical objects corresponds to creating a new group node, and then moving the two objects to be children of the new group node.
\end{itemize}
As these operations are so common, it is not obvious why a move operation should be difficult in a replicated setting.
In this section we demonstrate by example some of the problems that arise with replicated trees, before proceeding to our solution in \S\ref{sec:algorithm}.

@@

Another possible resolution is shown in Figure~\ref{fig:move-cycle}b: the nodes involved in the concurrent moves (and their children) could be duplicated, so that both ``$A$ as a child of $B$'' and ``$B$ as a child of $A$'' can exist in the tree.
However, such duplication is undesirable for the same reasons as in \S\ref{sec:move-same-item}.

In our opinion, the best way of handling the conflicting operations of Figure~\ref{fig:move-cycle} is to choose either Figure~\ref{fig:move-cycle}c or~\ref{fig:move-cycle}d: that is, either the result of applying replica 1's operation and ignoring replica 2's operation, or vice versa.
Like in \S\ref{sec:move-same-item}, the winning operation can be picked based on an operation timestamp.

@@

Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} have previously implemented a replicated filesystem with a move operation, and analysed the case of concurrent move operations introducing a cycle, as in \S\ref{sec:move-cycle}.
Using the CISE ('Cause I'm Strong Enough) proof tool~\cite{DBLP:conf/popl/GotsmanYFNS16,Najafzadeh:2016fi} the authors confirm that it is not sufficient for the replica that generates a move operation to check whether the operation introduces a cycle: like in Figure~\ref{fig:move-cycle}, two concurrent operations may be safe individually, but introduce a cycle when combined.

@@

While these solutions are valid, the authors go on to claim that ``no file system can support an unsynchronised move without anomalies, such as loss or duplication''~\cite{Najafzadeh:2018bw}.
We revisit that claim in this paper: our algorithm does not perform any locking, coordination or synchronisation among replicas, but it nevertheless ensures that the tree invariants are always satisified (in particular, it never introduces cycles), and it never duplicates or loses any tree nodes.
To our knowledge, our algorithm is the first to provide all of these properties simultaneously.
We give a precise specification of our algorithm's consistency properties in \S\ref{sec:proof}.

@@

The key consistency property of our algorithm is \emph{convergence}: that is, whenever any two replicas have applied the same set of operations, then they must be in the same state---even if the operations were applied in a different order on different replicas.
We prove this in \S\ref{sec:proof} by showing that applying operations is commutative.

Figure~\ref{fig:code} gives the full source code for our algorithm, implemented using the Isabelle/HOL language~\cite{DBLP:conf/tphol/WenzelPN08}.
We choose this language because it combines the conciseness of pseudocode with the precision of mathematical notation.
It supports formal reasoning, allowing us to prove the correctness of the algorithm (\S\ref{sec:proof}), and it also allows executable code to be exported to common functional programming languages such as Scala, Haskell, or OCaml.

@@

Recall from \S\ref{sec:move-cycle} that in order to prevent a cycle being introduced, the node being moved must not be an ancestor of the destination node.
To implement this in Figure~\ref{fig:code}, we first define the \isa{ancestor} relation in lines 22--24.
This relation is the transitive closure of the parent-child relation of a tree: if \isa{(p, m, c) {\isasymin} tree} then \isa{p} is an ancestor of \isa{c} (line 23); moreover, if \isa{(p, m, c) {\isasymin} tree} and \isa{a} is an ancestor of \isa{p}, then \isa{a} is also an ancestor of \isa{c} (line 24).
The \textbf{inductive} keyword indicates that this recursive definition is iterated until the least fixed point is reached.

The \isa{do\_op} function (lines 26--31) now performs the actual work of applying a move operation.
This function takes as argument a pair consisting of a \isa{Move} operation and the current tree, and it returns a pair consisting of a \isa{LogMove} operation (which will be added to the log) and an updated tree.
In line 29, the \isa{LogMove} record is constructed as described in \S\ref{sec:state-log}, obtaining the prior parent and metadata of \isa{c} using the \isa{get\_parent} function.

@@

Due to the undo-do-redo cycle, the state of the tree is as if all operations had been applied using \isa{do\_op} in increasing timestamp order, regardless of the order in which they were actually applied.
This provides a clear approach to the handling of conflicts:
\begin{itemize}
    \item If two operations concurrently move the same node, then the ``winning'' operation is the one with the higher timestamp.
        Effectively, the operation with the lower timestamp moves the node first, and then the operation with the higher timestamp moves it again.
        Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
    \item If two operations would introduce a cycle when combined, as in \S\ref{sec:move-cycle}, then the operation with the greater timestamp is ignored.
        This is the case because \isa{do\_op} checks for cycles based on the tree created by all operations with a lower timestamp.
        The lower of two conflicting operations will take effect, since that operation by itself is safe.
        When the higher-timestamped operation is applied, \isa{do\_op} detects that it would introduce a cycle, and therefore ignores the operation.
\end{itemize}

@@

In a distributed filesystem, replication and conflict resolution is required not only for the directory structure, but also for the contents of individual files.
This can be accomplished by using CRDTs for file contents as well.
For example, we have explored CRDT algorithms for applications using a JSON data model~\cite{Kleppmann:2016ve,LocalFirst} and implemented them in a library called Automerge~\cite{Automerge}.
We discuss distributed filesystems further in \S\ref{sec:filesystems}.

@@

We now discuss the correctness properties of the algorithm from \S\ref{sec:algorithm}.
All theorems stated here have been formally proved and mechanically checked using Isabelle.
For space reasons we give only a brief sketch of the proofs in this paper, but the full details can be found in our open source Isabelle files~\cite{SourceFiles}.

To reason about the state of a replica we define the function \isa{apply\_ops} on lines 53--54 of Figure~\ref{fig:code}.
It takes a list of operations \isa{ops} and returns the state of a replica after it has applied all the operations in \isa{ops}.
The \isa{apply\_ops} function works by starting in the initial state \isa{([], \{\})} consisting of the empty operation log \isa{[]} and the tree represented by the empty set \isa{\{\}}, and then applying the operations one by one to the state using the \isa{apply\_op} function (introduced in \S\ref{sec:applying}).
The \isa{foldl} function from the Isabelle/HOL standard library performs the iteration over the list of operations.

@@

In fact, we slightly generalise this property: we allow more than one root to exist, so the graph is actually a forest.
This allows an application to move nodes between multiple trees, if desired.
For example, the ``trash'' node used for deletion (\S\ref{sec:ops-trees}) can be separate from the main tree.

@@

As discussed in \S\ref{sec:applying}, we require that when replicas apply the same set of operations, they converge towards the same state, regardless of the order in which the operations are applied.
We formalise this in Isabelle as follows:
\begin{isabelle}
\isacommand{theorem}\isamarkupfalse%
\ apply{\isacharunderscore}ops{\isacharunderscore}commutes{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ {\isacartoucheopen}set\ ops{\isadigit{1}}\ {\isacharequal}\ set\ ops{\isadigit{2}}{\isacartoucheclose}\isanewline
\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{1}}{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \ \ \isakeyword{and}\ {\isacartoucheopen}distinct\ {\isacharparenleft}map\ move{\isacharunderscore}time\ ops{\isadigit{2}}{\isacharparenright}{\isacartoucheclose}\isanewline
\ \ \isakeyword{shows}\ {\isacartoucheopen}apply{\isacharunderscore}ops\ ops{\isadigit{1}}\ {\isacharequal}\ apply{\isacharunderscore}ops\ ops{\isadigit{2}}{\isacartoucheclose}
\end{isabelle}

@@

Isabelle/HOL can generate executable Haskell, OCaml, Scala, and Standard ML code from HOL definitions using a sophisticated code generation mechanism~\cite{DBLP:conf/flops/HaftmannN10}.
However, not all definitions can be realised in executable form: for example, the use of choice principles (like in \isa{get\_parent}) and inductively defined relations (e.g.\ \isa{ancestor}) cause problems.
Moreover, HOL's \isa{set} type allows infinite sets.
Whilst these constructs are convenient for theorem proving, they do not translate well to efficient code.
We therefore produce variants of the definitions in Figure~\ref{fig:code} that are designed for execution rather than theorem proving.
Rather than representing the tree as a set of (parent, metadata, child) triples, these definitions use a hash-map in which the keys are child nodes, and the values are (metadata, parent) pairs, written in Isabelle as \isa{('n::\{hashable\}, 'm {\isasymtimes} 'n) hm}.
The \isa{hashable} type class on keys means that keys must have an associated hashing function.
In effect, this hash-map is an index over the set of triples, using the fact that the parent and metadata for a given child are unique (\S\ref{sec:unique-parent}).
We use the hash-map implementation from the Isabelle Collections Framework~\cite{DBLP:conf/itp/LammichL10}.

@@

That is, if both algorithms are applied to the same operations then they produce identical logs and trees that contain the same set of bindings---i.e. the two trees are extensionally equivalent, despite having in-memory representations.
Moreover, we have corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic}, respectively, further establishing the correctness of our executable algorithms:\footnote{The ``executable'' analogue of \isa{apply\_ops\_unique\_parent} is trivially true, as the output tree is stored as a hash-map, and therefore too trivial to state as a theorem.}:

@@

Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
CRDTs have been defined for various abstract datatypes, such as registers and counters \cite{Shapiro:2011wy,Shapiro:2011un}, maps \cite{Baquero:2016iv,Kleppmann:2016ve}, sets \cite{Bieniusa:2012wu,Bieniusa:2012gt}, lists (sequences) \cite{Roh:2011dw,Nedelec:2013ky}, and text \cite{Preguica:2009fz,Weiss:2010hx}.
For tree data structures, several approaches have been proposed:
\begin{itemize}
    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support a move operation.
    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two alternative implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} survey approaches to handling conflicts on trees without providing concrete algorithms.
    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
        Some conflicts are handled by duplicating tree nodes.
        Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in \S\ref{sec:move-is-hard}.
    \item In our previous unpublished work \cite{ExtendedVersion} we develop the foundations for the algorithm described in this paper.
\end{itemize}
